---
title: "Bike use and bike and microvehicles crashes over time"
date: 
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(httr)
library(purrr)
library(plotly)
library(patchwork)

options(scipen=999)

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```


```{r filter function, include = FALSE}

filter_for = function(dataset, type) {
  if (type == "micro") {
  dataset %>% filter(str_detect(vehicle_options, "[Bb]ike") | 
           str_detect(vehicle_options, "REVEL") | 
           str_detect(vehicle_options, "SCO")  |
           str_detect(vehicle_options, "MOP")   |
           str_detect(vehicle_options, "ELEC")  |
           str_detect(vehicle_options, "^E-")) %>% 
  filter(vehicle_options != "ESCOVATOR" & vehicle_options != "Bike" &
      str_detect(vehicle_options, "Dirt", negate = TRUE),
      str_detect(vehicle_options, "[Mm]otorbike", negate = TRUE)
           )
}

  else if (type == "bike") {
  dataset %>% filter(str_starts(vehicle_options, "[Bb]ike")) 
  }
}
```

```{r load bike crash data, include = FALSE}
bike_crash_agg = 
  read_csv("./data/crash_dat.csv") %>%
  filter_for("bike") %>% 
  group_by(date) %>% 
  summarize(daily_bikes_crashed = n())

# This will give us the daily total of bikes that were involved in crashes, rather
# than the daily total of crashes that involved bikes. ie, if there was one crash 
# involving two bikes, that will count as two. (as explained in first version in import file)
```

```{r load microvehicle data, include = FALSE}
microvehicle_crash_agg = 
  read_csv("./data/crash_dat.csv") %>%
  filter_for("micro") %>% 
  group_by(date) %>% 
  summarize(daily_microveh_crashed = n()) 

# This will give us the daily total of microvehicles that were involved in crashes, rather
# than the daily total of crashes that involved microvehicles. ie, if there was one crash 
# involving two bikes, that will count as two. (as explained in first version in import file)
```

```{r load bike count data, include = FALSE}
bike_count_agg = 
  read_csv("./data/bike_count_df.csv")
```

On this page, we'll explore trends of bicycle use and bicycle and motorvehicle crashes over time, from 2017 through the end of October of 2020. This birds-eye-view will let us start to get a sense of whether there were noticeable differences pre and post covid, and of general trends over time. 

## Bike and microvehicle use and crashes in NYC, 2017-2020

In order to get a sense of bicycle use and bicycle and microvehicle crashes in the context of the covid pandemic in NYC, we start by looking at bicycle and microvehicle crashes and bicycle use over time. The plots below show, from top to bottom:
* Daily number of bicycles counted in NYC by the Department of Transportation's counters from 2017-2020
* Daily number of bicycles involved in crashes in NYC according to NYPD data on motorvehicle collisions from 2017-2020
* Daily number of microvehicles (including scooters, mopeds, electric bikes, e-bikes, and revel scooters) involved in crashes in NYC according to NYPD data on motorvehicle collisions from 2017-2020. 

Note that crashes are quantified as the number of bicycles (or microvehicles) involved, rather than the number of crashes - ie, if one crash occured involving two bicycles, that would count as two.

```{r initial plots over time, collapse = TRUE}
microveh_crash_plot = microvehicle_crash_agg %>% 
  ggplot(aes(x = date, y = daily_microveh_crashed)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Microvehicles involved in crashes in NYC daily: 2017-2020",
    x = "Date",
    y = ""
  )

bike_crash_plot = bike_crash_agg %>% 
  ggplot(aes(x = date, y = daily_bikes_crashed)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Bicycles involved in crashes in NYC daily: 2017-2020",
    x = "Date",
    y = ""
  )
  
bike_count_plot = bike_count_agg %>%
  ggplot(aes(x = date, y = total_daily_bikes)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Bicycle use in NYC: Bicycles counted daily by NYC DOT 2017-2020",
    x = "Date",
    y = ""
  )

bike_count_plot / bike_crash_plot / microveh_crash_plot
```

As we can see, there's a strong seasonal trend for bicycle use and a corresponding seasonal trend for bicycles involved in crashes. 

## Overlay years

**Bike use**
In order to take a more detailed look at how bicycle use and may have differed during 2020, we'll overlay each year of daily bicycle counts onto one plot. Click on the years to select/unselect.

```{r bike use overlay, collapse = TRUE}
bike_count_yr_plot = bike_count_agg %>% 
  mutate(year = lubridate::year(date),
         month = lubridate::month(date),
         day_of_month = lubridate::mday(date), 
         common_date = lubridate::mdy(paste(month, day_of_month, "2020")),
         year = as.factor(year)) %>%  
  ggplot(aes(x = common_date, y = total_daily_bikes, 
             group = year, color = year)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) + 
  labs(
    title = "Bicycles counted daily by NYC DOT 2017-2020",
    x = "Month",
    y = "Number of Collisions"
    ) +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10)) +
  scale_colour_discrete("Year") +
  scale_x_date(date_breaks = "1 month", labels = function(x) format(x, "%b")) +
  scale_color_viridis_d(end = .8)

ggplotly(bike_count_yr_plot)
```

As we can see, 2020 started out with higher bike use than the previous years, dipped slightly, and then surpassed the other years seasonal bike counts starting around July. From this visualizaion, we can't say definitively if COVID played a role in trends related to bicycle use, but it is interesting to observe how trends in 2020 differed from prior years.

**Bike crashes**

We can visualize the bike crash data in a similar way.

```{r bike crash overlay, collapse = TRUE}
bike_crash_yr_plot = bike_crash_agg %>% 
    mutate(year = lubridate::year(date),
           month = lubridate::month(date),
           day_of_month = lubridate::mday(date), 
           common_date = lubridate::mdy(paste(month, day_of_month, "2020")),
           year = as.factor(year)) %>% 
  ggplot(aes(x = common_date, y = daily_bikes_crashed, 
             group = year, color = year)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) + 
  labs(
    title = "Bicycles involved in crashes daily in NYC 2017-2020",
    x = "Month",
    y = "Bikes involved in crashes"
    ) +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10)) +
  scale_colour_discrete("Year") +
  scale_x_date(date_breaks = "1 month", labels = function(x) format(x, "%b")) +
  scale_color_viridis_d(end = .8)

ggplotly(bike_crash_yr_plot)
```

Bike crash data in 2020 follows a similar curve to bike use data - however, it is noticeably offset lower than bike crash data from previous years. 

**Crashes per observed bike**

In order to more specifically visualize whether crashes in 2020 were fewer per bicyclist, we look at a similar visualization plotting bikes involved in crashes per counted bicycle. This gives a sense of how bike safety (as measured by likelihood of being involved in a crash) has evolved over time in NYC.

Note that, since bike count is based on bicycles that passed counters maintained by the DOT and represent only a sample of all bicycles used daily in NYC, these ratios should not be interpreted as estimates of the rate of crashing. Nevertheless, they can be used to illustrate trends over time.

```{r crashes per bike, collapse = TRUE}
bike_count_and_crash = 
  inner_join(
    bike_count_agg,
    bike_crash_agg,
    by = "date"
  ) %>% 
  mutate(crashes_per_bike = daily_bikes_crashed/total_daily_bikes)

bike_count_crash_yr_plot = bike_count_and_crash %>% 
    mutate(year = lubridate::year(date),
           month = lubridate::month(date),
           day_of_month = lubridate::mday(date), 
           common_date = lubridate::mdy(paste(month, day_of_month, "2020")),
           year = as.factor(year)) %>%
  ggplot(aes(x = common_date, y = crashes_per_bike, 
             group = year, color = year)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) + 
  labs(
    title = "Crashes per counted bike daily in NYC 2017-2020",
    x = "Month",
    y = "Crashes per bike"
    ) +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10)) +
  scale_colour_discrete("Year") +
  scale_x_date(date_breaks = "1 month", labels = function(x) format(x, "%b")) +
  scale_color_viridis_d(end = .8)

ggplotly(bike_count_crash_yr_plot)
```

Crashes per bicycle counted appear remain steady overall, and don't have a clear seasonal pattern. Crashes per bicycle dip starting in the second half of 2019 through the first half of 2020, when they rise again to reflect trends in previous years.

**Microvehicle crashes**

Finally, while we don't have data on microvehicle use, we can look at overlaid microvehicle crashes from 2017-2020 in a similar way.


