---
title: "Visualizations and Mapping"
date: "11/22/2020"
output: 
  html_document:
  always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(httr)
library(purrr)
library(plotly)
library(jsonlite)
library(knitr)
options(scipen = 999)

Sys.setenv('MAPBOX_TOKEN' = 'pk.eyJ1IjoiYWFicmFtb3Y5MCIsImEiOiJja2gyZm5obzQwNWIxMnFxc3phcWh1MWtwIn0.amAvJHtFkTl9XWJ68fh96Q')

theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```


```{r import MV crash, include=FALSE}
#Importing crash data using the NYC Open Data API. The API response is limited to 50,000 rows, so we need a function that can be used to page through these results to capture all crashes in 2020. 

crash_api = function(offset, limit = 50000) {
  GET("https://data.cityofnewyork.us/resource/h9gi-nx95.csv", 
      query = list("$where" = "crash_date between '2019-01-01T00:00:0' and '2020-10-31T12:00:00'", "$limit" = limit, "$offset" = offset)) %>% 
  content("parsed") %>%
  as_tibble() 
}
offsets = seq(25000, 275000, by = 50000)
crash_dat = bind_rows(crash_api(offset = 0, limit = 25000),
                    map_df(offsets, crash_api)) 
crash_dat = crash_dat %>% distinct
```


```{r, tidy MV crash}
#cleaning - transpose so vehicle types are listed in one column
crash_dat2 = 
  crash_dat %>% 
  mutate(
    dow = as.factor(weekdays(crash_date))
  ) %>%
  separate(crash_date, into = c("year", "month", "day"), sep = "-") %>%
  mutate(year = as.integer(year),
         month = as.integer(month),
         day = as.integer(day)) %>%
  pivot_longer(
    vehicle_type_code1:vehicle_type_code_5,
    names_to = "vehicle_number",
    values_to = "vehicle_options",
    names_prefix = "vehicle_type_code"
   ) %>%
  drop_na(vehicle_options)
```


Subset to crashes involving a bike

```{r bikes}
bikes = crash_dat2 %>%
  filter(str_detect(vehicle_options, "[Bb]ike")
           ) 
           
#Run to check what the filter ^ resulted in
bikes %>%
  group_by(vehicle_options) %>%
  count() %>% View()
```




Exploring contributing factors for crashes that result in an injury to a cyclist. Subsetting to crashes in March - October and classifying 2019 as pre-COVID and 2020 as COVID.

```{r contributing_factor}
crash_factor = 
  bikes %>%
  filter(
    month %in% (3:10),
    number_of_cyclist_injured > 0
    ) %>%
  pivot_longer(
    contributing_factor_vehicle_1:contributing_factor_vehicle_5,
    names_to = "vehicle_factor_num",
    names_prefix = "contributing_factor_vehicle_",
    values_to = "contributing_factor"
   ) %>%
  mutate(
    covid = as.factor(if_else(year == 2019, 'Pre-COVID', 'COVID'))
  ) %>%
  group_by(covid, contributing_factor) %>%
  summarise(
    collisions = n_distinct(collision_id)
  ) %>%
  drop_na(contributing_factor) %>%
  ungroup(contributing_factor) %>%
  filter(contributing_factor != "Unspecified") %>%
  mutate(
    covid = fct_relevel(covid, 'Pre-COVID'),
    proportion = round(collisions / sum(collisions), 2) 
  ) 
  

top_10 = 
  crash_factor %>%
  group_by(covid) %>%
  top_n(n = 10, wt = proportion) %>%
  mutate(
    contributing_factor = fct_reorder(as.factor(contributing_factor), collisions, .desc = TRUE)
  ) %>%
  ggplot(aes(x = contributing_factor, y = proportion, fill = covid)) +
  geom_bar(stat = "identity") +
  facet_grid( . ~covid, scales = "free_x", space = "free_x") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), legend.position = "none") +
  labs(
    title = "Top Ten Contributing Factors for Crashes Involving Injuries to Cyclists",
    x = "",
    y = "Proportion of Crashes"
  )

ggplotly(top_10) 

```

The leading contributing causes for cyclist injuries did not change during COVID, with the majority of crashes involving driver error, such as driver inattention and failure to yield. There is a slight decrease in the proportion of crashes involving cyclist error.  


Plotting a heat map of bike crashes by time of day and day of week pre and during COVID.

```{r heat_plots}
crash_day_time = 
  bikes %>%
  filter(
    month %in% (3:10),
    number_of_cyclist_injured > 0
    ) %>%
  separate(crash_time, into = c("hour", "minute", "second"), sep = ":") %>%
  mutate(
    covid = as.factor(if_else(year == 2019, 'Pre-COVID', 'COVID'))
  ) %>%
  group_by(covid, hour, dow) %>%
  summarise(
    collisions = n_distinct(collision_id)
  ) %>%
  mutate(
    covid = fct_relevel(covid, 'Pre-COVID'),
    dow = fct_relevel(dow, "Sunday", "Saturday", "Friday", "Thursday", "Wednesday", "Tuesday")
  ) 


a <- list(
  text = "Pre-COVID",
  xref = "paper",
  yref = "paper",
  yanchor = "bottom",
  xanchor = "center",
  align = "center",
  x = 0.5,
  y = 1,
  showarrow = FALSE
)

b <- list(
  text = "COVID",
  xref = "paper",
  yref = "paper",
  yanchor = "bottom",
  xanchor = "center",
  align = "center",
  x = 0.5,
  y = 1,
  showarrow = FALSE
)


pre_heat = 
  crash_day_time %>%
  filter(covid == "Pre-COVID") %>%
  plot_ly(
    x = ~hour, y = ~dow, z = ~collisions, type = "heatmap", colors = "BuPu"
  ) %>%
  hide_colorbar() %>%
  layout(annotations = a)

covid_heat = 
  crash_day_time %>%
  filter(covid == "COVID") %>%
  plot_ly(
    x = ~hour, y = ~dow, z = ~collisions, type = "heatmap", colors = "BuPu"
  ) %>%
  hide_colorbar() %>%
  layout(annotations = b)

subplot(pre_heat, covid_heat, nrows = 2, margin = .1)

```


Overall, there’s a similar trend pre and during COVID where ethe majority of bike crashes are clustered in the evening. There seems to be slightly more clustering in the evening commute hours (4-6 PM) during COVID than pre-COVID and even fewer crashes in the morning hours during COVID.  


Replicating these heat maps but subsetting to crashes that have a contributing factor attributed to the cyclist (either a factor of cyclist error or a factor with the same vehicle number as the bike vehicle number).  

```{r heat_plots_cyclist}
day_time_bike_err = 
  bikes %>%
  filter(
    month %in% (3:10),
    number_of_cyclist_injured > 0
    ) %>%
  pivot_longer(
    contributing_factor_vehicle_1:contributing_factor_vehicle_5,
    names_to = "vehicle_factor_num",
    names_prefix = "contributing_factor_vehicle_",
    values_to = "contributing_factor"
   ) %>%
  filter(contributing_factor == "Pedestrian/Bicyclist/Other Pedestrian Error/Confusion" |
           vehicle_factor_num == vehicle_number) %>%
  separate(crash_time, into = c("hour", "minute", "second"), sep = ":") %>%
  mutate(
    covid = as.factor(if_else(year == 2019, 'Pre-COVID', 'COVID'))
  ) %>%
  group_by(covid, hour, dow) %>%
  summarise(
    collisions = n_distinct(collision_id)
  ) %>%
  mutate(
    covid = fct_relevel(covid, 'Pre-COVID'),
    dow = fct_relevel(dow, "Sunday", "Saturday", "Friday", "Thursday", "Wednesday", "Tuesday")
  ) 

pre_bike_err = 
  day_time_bike_err %>%
  filter(covid == "Pre-COVID") %>%
  plot_ly(
    x = ~hour, y = ~dow, z = ~collisions, type = "heatmap", colors = "Blues"
  ) %>%
  hide_colorbar() %>%
  layout(annotations = a)

covid_bike_err = 
  day_time_bike_err %>%
  filter(covid == "COVID") %>%
  plot_ly(
    x = ~hour, y = ~dow, z = ~collisions, type = "heatmap", colors = "Blues"
  ) %>%
  hide_colorbar() %>%
  layout(annotations = b)

subplot(pre_bike_err, covid_bike_err, nrows = 2, margin = .1)

```


Similar to heat maps above, crashes where the cyclist made an error are clustered in the evening hours and there is slightly more clustering during COVID than pre-COVID.

Again replicating these heat maps but subsetting to crashes that have a contributing factor attributed to the driver. 

```{r heat_plots_driver}
day_time_drive_err = 
  bikes %>%
  filter(
    month %in% (3:10),
    number_of_cyclist_injured > 0
    ) %>%
  pivot_longer(
    contributing_factor_vehicle_1:contributing_factor_vehicle_5,
    names_to = "vehicle_factor_num",
    names_prefix = "contributing_factor_vehicle_",
    values_to = "contributing_factor"
   ) %>%
  filter(
    contributing_factor != "Pedestrian/Bicyclist/Other Pedestrian Error/Confusion",
    vehicle_factor_num != vehicle_number
    ) %>%
  separate(crash_time, into = c("hour", "minute", "second"), sep = ":") %>%
  mutate(
    covid = as.factor(if_else(year == 2019, 'Pre-COVID', 'COVID'))
  ) %>%
  group_by(covid, hour, dow) %>%
  summarise(
    collisions = n_distinct(collision_id)
  ) %>%
  mutate(
    covid = fct_relevel(covid, 'Pre-COVID'),
    dow = fct_relevel(dow, "Sunday", "Saturday", "Friday", "Thursday", "Wednesday", "Tuesday")
  ) 


pre_drive_err = 
  day_time_drive_err %>%
  filter(covid == "Pre-COVID") %>%
  plot_ly(
    x = ~hour, y = ~dow, z = ~collisions, type = "heatmap", colors = "YlGn"
  ) %>%
  hide_colorbar() %>%
  layout(annotations = a)

covid_drive_err = 
  day_time_drive_err %>%
  filter(covid == "COVID") %>%
  plot_ly(
    x = ~hour, y = ~dow, z = ~collisions, type = "heatmap", colors = "YlGn"
  ) %>%
  hide_colorbar() %>% 
  layout(annotations = b)

subplot(pre_drive_err, covid_drive_err, nrows = 2, margin = .1)

```

Similar to heat maps above, crashes where the driver made an error are clustered in the evening hours and there is slightly more clustering during COVID than pre-COVID. 